Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-08-03 paper.tex

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. However, if the boost from Sommerfeld
enhancement (SFE) is small then the gamma-ray emission induced by CRs
is dominating most clusters with the exception of a few clusters; most
notably Fornax that shows the largest ratio of DM annihilation-to-CR
flux. In particular, we identify Fornax and Virgo as prime targets for
DM observations, Perseus for CR induced emission and we additionally
decide in favor of the well studied cluster Coma for comparison. While
studying these four clusters in great detail, we complement the
analysis by calculating the expected gamma-ray fluxes of a nearby
X-ray flux complete sample of galaxy clusters from the Rosat all-sky
survey (extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.} For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

@@

As expected, the analytical CR model accounts for the ``scatter'' to
the scaling relation and biases the gamma-ray flux high for prominent
cool core clusters by up to a factor of 10 relative to the expectation
from the scaling relation. This is due to the high target gas
densities and associated CR densities in those cool cores that shape
the spatial emission characteristic to be very similar to that
observed in thermal X-rays. We caution that these predictions only
apply for the case of negligible active CR transport such as CR
streaming and diffusion relative to the gas. Gamma-ray fluxes might be
considerably smaller in those cluster that do not show an extended
diffuse radio(-mini) halo that might point to a centrally concentrated
CR distribution \citep{2011A&A...527A..99E}. The CR spectrum $E^2
dN/dE$ shows the characteristic pion bump at energies around 1~GeV. It
is expected to dominate the DM annihilation signal of BM models for
clusters/groups, although we find that the gamma-ray flux induced by
the BM is higher for about 1/5 of the clusters in the HIFLUGCS
catalogue, where Fornax and M49 have the highest signal to noise. We
find that \Fermi-LAT will be able to start constraining these DM
models within five years using clusters of galaxies.

Combining extended \Fermi upper limits and our model flux predictions,
we limit the relative CR pressure, $X_\CR = P_\CR/P_\rmn{th}$, in 32
nearby bright galaxy clusters of the \Fermi sample. The best limits
are found in Norma and Coma of the order of 3\%, with typical limits
around 10\%. This is comparable to those inferred by the \Fermi
collaboration \citep{2010ApJ...717L..71A} and mainly due to the
differently assumed Bayesian prior on the CR pressure which implies a
different cluster mass dependence of the resulting limits on $X_\CR$.

>>>>

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. However, if the boost from Sommerfeld
enhancement (SFE) is small then the gamma-ray emission induced by CRs
is dominating most clusters with the exception of a few clusters; most
notably Fornax that shows the largest ratio of DM annihilation-to-CR
flux. In particular, we identify Fornax and Virgo as prime targets for
DM observations, Perseus for CR induced emission and we additionally
decide in favor of the well studied cluster Coma for comparison. While
studying these four clusters in great detail, we complement the
analysis by calculating the expected gamma-ray fluxes of a nearby
X-ray flux complete sample of galaxy clusters from the Rosat all-sky
survey (extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.} For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

@@

As expected, the analytical CR model accounts for the ``scatter'' to
the scaling relation and biases the gamma-ray flux high for prominent
cool core clusters by up to a factor of 10 relative to the expectation
from the scaling relation. This is due to the high target gas
densities and associated CR densities in those cool cores that shape
the spatial emission characteristic to be very similar to that
observed in thermal X-rays. We caution that these predictions only
apply for the case of negligible active CR transport such as CR
streaming and diffusion relative to the gas. Gamma-ray fluxes might be
considerably smaller in those cluster that do not show an extended
diffuse radio(-mini) halo that might point to a centrally concentrated
CR distribution \cite{2011A&A...527A..99E}. The CR spectrum $E^2
dN/dE$ shows the characteristic pion bump at energies around 1~GeV. It
is expected to dominate the DM annihilation signal of BM models for
clusters/groups, although we find that the gamma-ray flux induced by
the BM is higher for about 1/5 of the clusters in the HIFLUGCS
catalogue, where Fornax and M49 have the highest signal to noise. We
find that \Fermi-LAT will be able to start constraining these DM
models within five years using clusters of galaxies.

Combining extended \Fermi upper limits and our model flux predictions,
we limit the relative CR pressure, $X_\CR = P_\CR/P_\rmn{th}$, in 32
nearby bright galaxy clusters of the \Fermi sample. The best limits
are found in Norma and Coma of the order of 3\%, with typical limits
around 10\%. This is comparable to those inferred by the \Fermi
collaboration \cite{2010ApJ...717L..71A} and mainly due to the
differently assumed Bayesian prior on the CR pressure which implies a
different cluster mass dependence of the resulting limits on $X_\CR$.
